Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-18 writing/paper-rbm/paper.Rmd
include new files

@@

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that many data processes of interest are realistically not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

```{r sample-models, fig.show='hold', fig.height=2, fig.cap='Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.', out.width='100%'}
load("../resources/data/params_theta.Rdata")
params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
params_bad <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

@@

In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. The degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the standard overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. This work aligns with the call to action by Google scientists in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most commonly considered. In this paper, we consider the binary case for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \@ref(fig:rbm) with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification. For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in a RBM model [@hinton2006fast].

@@

Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \@ref(explorations-of-model-properties-through-simulation) that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
\begin{align}
(\#eq:pmf)
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}
where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities \@ref(eq:pmf) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, (\#eq:t)
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \@ref(eq:pmf) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \@ref(eq:pmf) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. In model fitting, a kind of Gibbs sampling can be tried due to the conditional architecture of the RBM (i.e. visibles given hiddens or vice verse). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \@ref(model-fitting)).

@@

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \@ref(eq:t), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a log-ratio of extremal
probabilities (LREP) of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) (\#eq:elpr)
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \@ref(eq:elpr) needs to stay bounded for a sequence of RBM models to be stable. That is, we make the following convention.
```{definition, name = "S-unstable RBM", echo=TRUE}
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
```
In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

>>>>

---
title: "Properties and Bayesian fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    shortname: Kaplan A.,
    numaddress: 1
    corresponding: true
  - name: Daniel Nordman
    shortname: D. Nordman, and
    corresponding: false
    numaddress: 2
  - name: Stephen Vardeman
    shortname: S. Vardeman
    numaddress: "2,3"
    corresponding: false
shortauthor: "KAPLAN \\textsc{et al}"
address:
  - numaddress: 1
    orgdiv: Department of Statistical Science
    orgname: Duke University
    orgstate: North Carolina
    orgcountry: USA
  - numaddress: 2
    orgdiv: Department of Statistics
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
  - numaddress: 3
    orgdiv: Department of Industrial and Manufacturing Systems Engineering
    orgname: Iowa State University
    orgstate: Iowa
    orgcountry: USA
corresponding: true
corresponding_name: Andee Kaplan
corresponding_address: Department of Statistical Science, Duke University, P.O. Box 90251, Durham, NC 27708-0251
corresponding_email: andrea.kaplan@duke.edu
output: 
  rmarkdown::pdf_document:
    keep_tex: true
    template: ../resources/latex/wileynjd-ams_article.tex
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
---

```{r setup, include=FALSE}
# https://github.com/rstudio/bookdown/pull/374
fix_envs <- function(x) {
  beg_reg <- '^\\s*\\\\begin\\{.*\\}'
  end_reg <- '^\\s*\\\\end\\{.*\\}'
  i3 = if (length(i1 <- grep(beg_reg, x))) (i1 - 1)[grepl("^\\s*$", x[i1 - 1])]
  i3 = c(i3, if (length(i2 <- grep(end_reg, x))) (i2 + 1)[grepl("^\\s*$", x[i2 + 1])])
  if (length(i3)) x = x[-i3]
  x
}

options(
  dplyr.print_min = 6, dplyr.print_max = 6, width = 70,
  digits = 3,
  bookdown.post.latex = fix_envs
)
knitr::opts_chunk$set(
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  echo = FALSE
)

ggplot2::theme_set(ggplot2::theme_bw(base_family="serif"))

library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(mvtnorm)
library(rootSolve)
library(agridat)
library(xtable)
```

@@

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that many data processes of interest are realistically not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \ref{fig:sample-models}. The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

```{r sample-models, fig.show='hold', fig.height=2, fig.cap='\\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.', out.width='100%'}
load("../resources/data/params_theta.Rdata")
params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
params_bad <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

@@

In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. The degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the standard overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \ref{explorations-of-model-properties-through-simulation} details our explorations into the model behavior and potential propriety issues with the RBM class. This work aligns with the call to action by Google scientists in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}. Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \ref{data-encoding}.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most commonly considered. In this paper, we consider the binary case for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification. For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in a RBM model [@hinton2006fast].

@@

Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}
where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities \eqref{eq:pmf} sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \eqref{eq:pmf} in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf} and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. In model fitting, a kind of Gibbs sampling can be tried due to the conditional architecture of the RBM (i.e. visibles given hiddens or vice verse). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$. Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \eqref{eq:t}, appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$. Here the mean of $\boldsymbol t(\boldsymbol x)$ is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a log-ratio of extremal
probabilities (LREP) of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be stable. That is, we make the following convention.

\begin{definition}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{definition}

In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
